### 2025-07-15 factor-analysis.qmd
20250715 - factor analysis and SEM

@@

A key tool when designing a structural equation model is a conceptual depiction of the hypothesized causal processes.
A path diagram depicts the hypothesized causal processes that link two or more variables.
@Karch2025_packages provides a tool to create `lavaan` `R` syntax from a path analytic diagram: <https://lavaangui.org>.

@@

Simple structure is achieved when each variable loads highly onto as few factors as possible (i.e., each item has only one significant or primary loading).\index{simple structure}
Oftentimes this is not the case, so we choose our rotation method in order to decide if the factors can be correlated (an oblique rotation) or if the factors will be uncorrelated (an orthogonal rotation).\index{factor analysis}\index{rotation}\index{orthogonal rotation}\index{oblique rotation}
If the factors are not correlated with each other, use an orthogonal rotation.\index{factor analysis}\index{rotation}\index{orthogonal rotation}
The correlation between an item and a factor is a factor loading, which is simply a way to ask how much a variable is correlated with the underlying factor.\index{factor analysis}\index{structural equation modeling!factor loading}
However, its interpretation is more complicated if there are correlated factors!\index{factor analysis}\index{rotation}\index{oblique rotation}

@@

: Criteria for Acceptable and Good Fit of Structural Equation Models Based on Fit Indices. {#tbl-semFitIndices}

>>>>

## Factor Analysis and Structural Equation Modeling

Factor analysis forms the measurement model component of a structural equation model (SEM).
The *measurement model* is what we settle on as the estimation of each construct before we add the structural component to estimate the relations among latent variables.
Basically, in a structural equation model, we add the structural component onto the measurement model.
For instance, our measurement model (i.e., based on factor analyis) might be to estimate, from a set of items, three latent factors: usage, aggressiveness, and performance.
Our structural model then may examine what processes (e.g., sport drink consumption or sleep) influence these latent factors, how the latent factors influence each other, and what the latent factors influence (e.g., fantasy points).
SEM is [confirmatory factor analysis](#sec-cfa) with regression paths that specify hypothesized causal relations between the latent variables (the structural component of the model).\index{latent variable}\index{structural equation modeling}\index{factor analysis!confirmatory}
Exploratory structural equation modeling (ESEM) is a form of SEM that allows for a combination of [exploratory factor analysis](#sec-efa) and [confirmatory factor analysis](#sec-cfa) to estimate latent variables and the relations between them.\index{latent variable}\index{structural equation modeling}\index{factor analysis!exploratory}\index{factor analysis!confirmatory}\index{structural equation modeling!exploratory}

@@

A key tool when designing a factor analysis or structural equation model is a conceptual depiction of the hypothesized causal processes.
A *path diagram* depicts the hypothesized causal processes that link two or more variables.
Path diagrams are an example of a causal diagram and are similar to [directed acyclic graphs](#sec-causalDiagrams) discussed in @sec-causalDiagrams.
@Karch2025_packages provides a tool to create `lavaan` `R` syntax from a path analytic diagram: <https://lavaangui.org>.

@@

Simple structure is achieved when each variable loads highly onto as few factors as possible (i.e., each item has only one significant or primary loading).\index{simple structure}
Oftentimes this is not the case, so we choose our rotation method in order to decide if the factors can be correlated (an oblique rotation) or if the factors will be uncorrelated (an orthogonal rotation).\index{factor analysis}\index{rotation}\index{orthogonal rotation}\index{oblique rotation}
If the factors are not correlated with each other, use an orthogonal rotation.\index{factor analysis}\index{rotation}\index{orthogonal rotation}
The correlation between an item and a factor is a factor loading, which is simply a way to ask how much a variable is correlated with the underlying factor.\index{factor analysis!factor loading}
However, its interpretation is more complicated if there are correlated factors!\index{factor analysis}\index{rotation}\index{oblique rotation}

@@

: Criteria for Acceptable and Good Fit of Factor Analysis and Structural Equation Models Based on Fit Indices. {#tbl-semFitIndices}
